---
title: "KWF_Beach_Sampling_Analysis_2019_2020"
output:
  html_document: 
    fig_width: 10
    fig_height: 6
    fig_caption: yes
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scroll: no
editor_options: 
  chunk_output_type: inline
---

Document last updated `r Sys.time()` by Benjamin Meyer ([ben\@kenaiwatershed.org](mailto:ben@kenaiwatershed.org){.email})

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

# clear environment
rm(list=ls())

# load packages
library(tidyverse)
library(lubridate)
library(readr)
library(readxl)
library(writexl)
library(hms)
library(plotly)
library(DT)
library(xlsx)
library(leaflet)
library(DT)
library(ggpubr)
library(plotrix)

select <- dplyr::select

# set plotting themes

## geom_col plots theme
col_theme <- theme(axis.title = element_text(size = 14, face = "bold"),
                   strip.text = element_text(size = 14, face = "bold"),
                   legend.title = element_text(size = 14, face = "bold"),
                   legend.text = element_text(size = 14),
                   axis.text = element_text(size = 14))

## geom_points plots theme
points_theme <- theme(axis.title = element_text(size = 14, face = "bold"),
                   strip.text = element_text(size = 14, face = "bold"),
                   legend.title = element_text(size = 14, face = "bold"),
                   legend.text = element_text(size = 14),
                   axis.text = element_text(size = 14),
                   title = element_text(size = 18))
```

<br>

This draft document contains preliminary data explorations of 2019-2020 beach bacteria sampling data from the Kenai River Watershed.

<br>

Notes:

-   Attempted to source beach sampling data from EPA water quality portal but did not find it there. Sourced data instead from AWQMS files and other files found on KWF server.

-   2020 MST data is present in AWQMS files, 2019 MST data is not. Sourced 2019 MST data from "MST results 2019.xlsx"

-   In the future, request that SGS can provide sample results in .csv or .xlsx format in addition to the PDFs to eliminate potential for data entry or data management error.

<br>

------------------------------------------------------------------------

### Read in data

```{r}

# import known data from local server source

# specify column import names
vars <- as.character(c("Monitoring Location Name",
            "Activity Latitude",
            "Activity Longitude",
            "Activity Medium",
            "Activity Start Date",
            "Activity Start Time",
            "Characteristic Name",
            "Result Analytical Method",
            "Result Value",
            "Result Value Units",
            "Result Qualifier"))

# specify new column names
new_vars <- as.character(c("location",
            "lat",
            "long",
            "medium",
            "date",
            "time",
            "data_type",
            "analytical_method",
            "val",
            "units",
            "qualifier"))


# import 2019 AWQMS data
## used this excel file because it had the most recent modified date
## see notes in excel file; not sure if all available data is contained within here
bs19 <- read_excel("data/KWF_2019_Kenai-BEACH_AWQMS_forQAQC_MH101119.xlsx", sheet = "AWQMS Data Template", skip = 7) %>%
  select(all_of(vars))

# import 2020 AWQMS data
## see notes in excel file; not sure if all avilable data is contained within here
bs20 <- read_excel("data/2020_KWF_Kenai-BEACH_AWQMS_SUBMITTED.xlsx", sheet = "RAW AWQMS Data Template", skip = 3) %>%
  select(all_of(vars))

# combine both years
dat <- bind_rows(bs19,bs20)

# rename columns
colnames(dat) <- new_vars

## 2019 MST data absent from AWQMS file.  So, import separately and format to join
## imported sheet was created manually on 12/8/20 from the sheet titled "Summary"
mst19 <- read_excel("data/MST results 2019.xlsx", sheet = "AWQMS_summary_format") %>%
  select(-site_id)

# join 2019 MST data to overall data set
dat <- bind_rows(dat,mst19)

# tidy column formats
dat <- dat %>%
  transform(time = hms::as_hms(time))

# replace "Gull_Gull-#" with "Gull_Gull-4"
# correct location names
dat <- dat %>%
  mutate(analytical_method = str_replace_all(analytical_method,c("Gull_Gull-5" = "Gull_Gull-4",
                                                                 "Gull_Gull-6" = "Gull_Gull-4"))) %>%
  mutate(location = str_replace_all(location, c("North Kenai River Beach 4" = "North Kenai Beach 4")))

rm(bs19,bs20,mst19)
                                             
```

<br>

Notes post data read-in

-   Air temperature values are all negative integers (maybe a code? do not use air temperature data from this data set)
-   Applied name correction to reconcile equivalent "North Kenai Beach 4" and "North Kenai River Beach 4" sites.
-   Some unit designations are missing (turbidity, weather condition, wind direction, others?)
-   Some Microbial Source Tracking analytic method ID is entered as "Gull_Gull-\#" (e.g. 5, 6, 7, etc) when the only designation for this method should be "Gull_Gull-4." Likely occurred from an excel drag-down at some point in data entry.
-   Many observations in "val" column are character (e.g. "MODERATE", "ND", etc). Caution when converting to numeric.
-   Verified that "NA" observations in the value column are actually absent, checked original PDF files from SGS.

<br>

Procedure:

-   Recreate tables (approximately) found in provided excel file ("2020 Beach Sampling Compiled Results_FINALREPORT.xlsx") to verify values
-   Create figures to visualize exceedence data

<br>

------------------------------------------------------------------------

### Microbial Source Tracking

#### MST Data Table

```{r}
# recreate table in tab "2019-2020_FinalReport_MST2020"
mst_tbl <- dat %>%
  filter(data_type == "Microbial Source Tracking") %>%
  arrange(location,date,analytical_method) %>%
  select(date,location,analytical_method,val) %>%
  pivot_wider(names_from = analytical_method, values_from = val) %>%
  arrange(date,location)

colnames(mst_tbl) <- c("Date","Location","Dog Feces","Gull Feces","Human Feces")

# print
datatable(mst_tbl,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))

```

<br>

#### MST Results Figure

```{r}
# prep table
mst_tbl <- dat %>%
  filter(data_type == "Microbial Source Tracking") %>%
  arrange(location,date,analytical_method) %>%
  select(date,location,analytical_method,val) %>%
  # replace results DNQ and ND with zero and convert to numeric
  mutate(val = as.numeric(str_replace_all(val,c("DNQ" = "0", "ND" = "0"))))

# plot
mst_tbl %>%
  ggplot(aes(as.factor(date),val, fill = analytical_method)) +
  geom_col(position = "dodge") +
  facet_grid(.~location, labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("Concentration (Copies/100 mL)") +
  theme_bw() +
   # modify legend
  scale_fill_discrete(name="Source",
                         breaks=c("Dog_BacCan-UCD", "Gull_Gull-4", "Human_HF183"),
                         labels=c("Dog", "Gull", "Human")) +
  # theme 
  theme(axis.text.x = element_text(vjust = .5, angle = 85),
        strip.text.y = element_text(angle = 0)) +
  col_theme

# save plot
ggsave("output/mst_kenai_beach_2019_2020.png", width = 12, height = 6)
```

<br>

------------------------------------------------------------------------

### Enterococci and Fecal Coliform

#### Bacteria Concentrations Data Table

```{r}
# prep table
bac_tbl <- dat %>% 
  filter(data_type %in% c("Enterococci","Fecal Coliform")) %>% 
  arrange(location,date,data_type) %>%
  select(date,location,data_type,val) %>%
  # replace results DNQ and ND with zero and convert to numeric
  mutate(val = as.numeric(str_replace_all(val,c("DNQ" = "0", "ND" = "0")))) %>%
  # use day of year for x axis and year for facet grid
  mutate(day = yday(date),
         year = year(date))

# print
datatable(bac_tbl,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))

```

<br>

#### Bacteria Concentrations Figures

```{r, fig.height=10, fig.width=12}
# plot 1
bac_tbl %>%
  ggplot(aes(x = as.Date(paste(2014,strftime(date,format="%m-%d"),sep="-")),
                           y = val, 
                           color = as.factor(year))) +
  geom_jitter(size = 3) +
  facet_grid(location~data_type, labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("Individual Sample Concentration (CFU/100 mL)") +
  scale_color_discrete(name = "Year")+
  # theme
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1),
        strip.text.y = element_text(angle = 0)) +
  points_theme

# save plot
ggsave("output/bacteria_conc_2019_2020_v1.png", width = 10, height = 10)

```

```{r}
# plot 2, enterococci
bac_tbl %>%
  filter(data_type == "Enterococci") %>%
  # note; use of year for date format is "fake" placeholder
  ggplot(aes(x = as.Date(paste(2014,strftime(date,format="%m-%d"),sep="-")),
                           y = val, 
                           color = as.factor(year))) +
  geom_jitter(size = 2) +
  facet_grid(. ~ location, scales = "free_y", labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("Individual Sample\nConcentration (CFU/100 mL)") +
  ylim(0,625) +
  ylim(0,625) +
  scale_color_discrete(name = "Year") +

  # single-sample threshold
  geom_hline(aes(yintercept = 130, linetype =  "Raw seafood consumption\n(130 CFU/100 mL)"), color = "blue") +
  scale_linetype_manual(name = "Threshold", values = c(2, 2),
                        guide = guide_legend(override.aes = list(color = c("blue")))) +
  # overall theme
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1,hjust = 1, angle = 35),
        strip.text.y = element_text(angle = 0)) +
  ggtitle("Enterococci Concentrations") +
  points_theme

# save plot
ggsave("output/enterococci_conc_2019_2020.png", width = 10, height = 4)


# plot 3, fecal coliform
bac_tbl %>%
  filter(data_type == "Fecal Coliform") %>%
  # note; use of year for date format is "fake" placeholder
  ggplot(aes(x = as.Date(paste(2014,strftime(date,format="%m-%d"),sep="-")),
                           y = val, 
                           color = as.factor(year))) +
  geom_jitter(size = 2) +
  facet_grid(. ~ location, scales = "free_y", labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("Individual Sample\nConcentration (CFU/100 mL)") +
  scale_color_discrete(name = "Year") +

  # single-sample thresholds
  geom_hline(aes(yintercept = 31, linetype =  "Raw Seafood Consumption\n(31 CFU/100 mL)"), color = "blue") +
  geom_hline(aes(yintercept = 400, linetype =  "Water recreation, secondary\n(400 CFU/100 mL)"), color = "red") +
  scale_linetype_manual(name = "Threshold", values = c(2, 2),
                        guide = guide_legend(override.aes = list(color = c("blue","red")))) +
  # overall theme
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1,hjust = 1, angle = 35),
        strip.text.y = element_text(angle = 0)) +
  points_theme +
  ggtitle("Fecal Coliform Concentrations")

# save plot
ggsave("output/fecal_coliform_conc_2019_2020.png", width = 10, height = 4)

```

<br>

------------------------------------------------------------------------

### Exeedence Frequency

-   What percentage of individual sample values exceed thresholds for each site?

-   What percentage of 30-day geometric mean values exceed thresholds for each site?

General Notes

-   Note: For geometric mean calculations, non-detect values were entered at 50% their method detection limit: See section 2.4.1 of ADEC "Listing Methodology for Determining Water Quality Impairments from Pathogens": "It is appropriate to consider non-detect samples when calculating the geometric mean for assessment purposes. Rather than eliminating the "non-detects" from the assessment data, these results and sample results measured below the detection limit will be calculated as 50% of the method detection limit. This approach may not be appropriate during the analysis of water quality trends."

Threshold values

-   Enterococci:

    -   "In a 30-day period, the geometric mean of samples may not exceed 35 enterococci CFU/100 mL, and not more than 10% of the samples may exceed a statistical threshold value (STV) of 130 CFU/100 mL."

-   Fecal coliform:

    -   "Harvest for consumption of raw mollusks or other raw seafood: The geometric mean of samples may not exceed 14 fecal coliform/100 ml; and not more than 10% of the samples may exceed 31 CFU/100 mL for a membrane filtration test."

    -   "Water recreation, secondary: In a 30-day period, the geometric mean of samples may not exceed 200 fecal coliform/100 mL, and not more than 10% of the samples may exceed 400 fecal coliform/100 mL."

<br>

Calculate 30-day rolling geometric mean values for Enterococci and Fecal Coliform

```{r}
library(zoo)
library(psych)

dat_gm30 <- dat %>%
  filter(data_type %in% c("Fecal Coliform","Enterococci")) %>%
  mutate(year = year(date)) %>%
  transform(val = as.numeric(val)) %>%
  # use 50% of value if it has an "Undetected" qualifier
  mutate(val = ifelse(is.na(qualifier),val,
                      ifelse(qualifier == "U",(val*.5),val))) %>%
  group_by(data_type,location,year) %>%
  arrange(year,data_type,location,date) %>%
  # calculate 30 day rolling geometric means
  mutate(gm30 = rollapplyr(val, 1:n() - findInterval(date - days(30), date), 
                           geometric.mean, fill = NA, partial = T)) %>%
  
  # mutate values from the first annual 30 days of sampling to NA. (We need 30 days for a 30-day rolling geometric mean value).  However, the sampling dates do not always line up so that the first 4 weeks of observations is exactly 30 days. Here, the rolling window is set to 25 days so as not to "throw away" that data. What we really want is a geometric mean of the first four weekly observations
  mutate(gm30 = as.numeric(ifelse(date < min(date + days(25)),"",gm30)))

```

<br>

Designate exceedence values

```{r}
# create columns of exceedence occurences
dat_excd <- dat_gm30 %>%
  # individual sample exceedences
  mutate(indv_excd = case_when(
    (val < 130 & data_type == "Enterococci") ~ "no_indv_excd",
    (val >= 130 & data_type == "Enterococci") ~ "ec_excd_130",
    (val < 31 & data_type == "Fecal Coliform") ~ "no_indv_excd",
    (val >= 31 & val < 400 & data_type == "Fecal Coliform") ~ "fc_excd_31",
    (val >= 400 & data_type == "Fecal Coliform") ~ "fc_excd_400")) %>%
  # 30-day rolling geometric mean value exceedences
  mutate(gm30_excd = case_when(
    (gm30 < 35 & data_type == "Enterococci") ~ "no_gm_excd",
    (gm30 >= 35 & data_type == "Enterococci") ~ "gm_ec_excd_35",
    (gm30 < 14 & data_type == "Fecal Coliform") ~ "no_gm_excd",
    (gm30 >= 14 & val < 200 & data_type == "Fecal Coliform") ~ "gm_fc_excd_14",
    (gm30 >= 200 & data_type == "Fecal Coliform") ~ "gm_fc_excd_200"))
```

Calculate % exceedence frequency by site, year, and bacteria type for individual samples

```{r}
  library(janitor)

# prop table for individual exceedences
indv_excd_tbl <- dat_excd %>%
  # remove observations with no value = NA
  filter(!is.na(val)) %>%
  select(location,data_type,year,indv_excd) %>%
  group_by(location,data_type,year) %>%
  count(indv_excd) %>%
  spread(key = indv_excd, val = n) %>%
  mutate_at(vars(contains("excd")),funs(as.numeric)) %>%
  mutate(across(everything(), ~replace_na(.x, 0))) %>%
  # sum across columns
  # was unable to find a way to successfully use a function to sum across columns here; used manual addition...
  mutate(sample_n = ec_excd_130 + fc_excd_31 + fc_excd_400 + no_indv_excd) %>%
  # calculate % of sample ct exceedence
  mutate(ec_excd_130 = ec_excd_130 / sample_n,
         fc_excd_31 = fc_excd_31 / sample_n,
         fc_excd_400 = fc_excd_400 / sample_n) %>%
  select(-no_indv_excd) 

```

<br>

#### Individual sample value exceedences

<br>

##### Table, overall frequency of individual sample exceedences

```{r}
indv_excd_tbl_exp <- indv_excd_tbl %>%
  select(location, data_type,year,sample_n,ec_excd_130,fc_excd_31,fc_excd_400) %>%
  transform(year = as.factor(year),
            sample_n = as.factor(sample_n)) %>%
  adorn_pct_formatting()

# prep table for export
colnames(indv_excd_tbl_exp) <- c("Location","Bacteria","Year","Sample Count",
                                 "% Enterococci Samples Exceeding 130 CFU/100 mL",
                                 "% Fecal Coliform Samples Exceeding 31 CFU/100 mL",
                                 "% Fecal Coliform Samples Exceeding 400 CFU/100 mL")
# save table in csv
write_csv(indv_excd_tbl_exp,"output/individual_sample_exceedences.csv")

# print table
datatable(indv_excd_tbl_exp,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))

```

<br>

##### Table, frequency of individual sample exceedences within 30-day periods

Recall; regarding individual sample values and 30-day periods:

Enterococci:

-   "... not more than 10% of the samples may exceed a statistical threshold value (STV) of 130 CFU/100 mL."

Fecal coliform:

-   "... not more than 10% of the samples may exceed 31 CFU/100 mL for a membrane filtration test."

-   "... not more than 10% of the samples may exceed 400 fecal coliform/100 mL."

```{r}
# calculate % of individual samples exceeding thresholds within 30-day periods
indv_excd_30day <- dat_excd %>%
  # designate 30-day periods for each site/year
  group_by(location,data_type,year) %>%
  mutate(period_30day = rep(seq_len(n()/2), each = 4, length.out = n()),
         sample_ct = n()) %>%
  ungroup() %>%
  # create columns of start/end dates for each 30-day period
  group_by(location,data_type,year,period_30day) %>%
  mutate(start_date = min(date),
         end_date = max(date)) %>%
  # count number of exceedences by type, site, year
  group_by(location,data_type,year,sample_ct,period_30day,start_date,end_date) %>%
  filter(!is.na(indv_excd)) %>%
  count(indv_excd) %>%
  # what % of individual samples exceeded thresholds during 30-day periods?
  mutate(pct_indv_samples_excd = (n / sample_ct)) %>%
  # remove instances of no individual sample exceedences
  filter(indv_excd != "no_indv_excd") 

# prep table for export
indv_excd_30day_exp <- indv_excd_30day %>%
  # retain only instances where > 10% of individual samples exceeded thresholds
  filter(pct_indv_samples_excd > .1)  %>%
  ungroup() %>%
  dplyr::mutate_all(as.character) %>%
  mutate(pct_indv_samples_excd = as.numeric(pct_indv_samples_excd)) %>%
  # adorn with percentage format
  adorn_pct_formatting() %>%
  select(location,data_type,year,indv_excd,start_date,end_date,pct_indv_samples_excd) %>%
  # replace exceedence type with longer names
  mutate(indv_excd = str_replace_all(indv_excd,c("fc_excd_31" = "Fecal Coliform > 31",
                                                                 "fc_excd_400" = "Fecal Coliform > 400",
                                                                 "ec_excd_130" = "Enterococci > 130"))) %>%
  spread(indv_excd,pct_indv_samples_excd) %>%
  arrange(year,location,data_type)

 # rename columns  
colnames(indv_excd_30day_exp) <- c("Location","Bacteria","Year","Start Date","End Date",
                                   "% Samples with Enterococci > 130 in 30-day period",
                                   "% Samples with Fecal Coliform > 31 in 30-day period")

# save table in csv
write.csv(indv_excd_30day_exp,"output/30day_periods_individual_sample_exceedences.csv", row.names = F)

# print table
datatable(indv_excd_30day_exp,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))

```

<br>

##### Figure, frequency of individual sample exceedences

```{r}
indv_excd_tbl %>%
  gather(excd, val, contains("excd")) %>%
  ggplot(aes(as.factor(year),val,fill = excd)) +
  geom_col(position = "dodge2") +
  facet_grid(. ~ location, labeller = labeller(location = label_wrap_gen(5))) +
  ylim(0,1) +
  xlab("") +
  ylab("% of Total Indvidual Samples\nExceeding Thresholds") + 
  theme_bw() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  # modify legend text
  scale_fill_discrete(name="Exceedence Type\n(CFU/100 mL)",
                      breaks = c("ec_excd_130","fc_excd_31","fc_excd_400"),
                      labels = c("Enterococci > 130",
                                 "Fecal Coliform > 31",
                                 "Fecal Coliform > 400")) +
  col_theme

# save image
ggsave("output/individual_sample_exceedences.png",width = 10, height = 4)

```

<br>

##### Figure, frequency of individual sample exceedences within unique 30-day periods

```{r}
indv_excd_30day %>%
  ggplot(aes(period_30day,pct_indv_samples_excd,fill = indv_excd)) +
  geom_col(position = position_dodge(preserve = "single")) +
  facet_grid(year ~ location, labeller = labeller(location = label_wrap_gen(5))) +
  ylim(0,.3) +
  xlab("30-Day Sampling Period") +
  ylab("% of Indvidual Samples\nExceeding Thresholds Within 30-Day Periods") + 
  theme_bw() +
  geom_hline(yintercept = .1, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  # modify legend text
  scale_fill_discrete(name="Exceedence Type\n(CFU/100 mL)",
                      breaks = c("ec_excd_130","fc_excd_31","fc_excd_400"),
                      labels = c("Enterococci > 130",
                                 "Fecal Coliform > 31",
                                 "Fecal Coliform > 400")) +
  col_theme

# save image
ggsave("output/30day_individual_sample_exceedences.png",width = 10, height = 4)


```

<br>

***
***

<br>

#### 30-day rolling mean of sample values, exceedences
```{r}

# plot, enterococci 30 day geometric means
dat_excd %>%
  filter(data_type == "Enterococci") %>%
  # note; use of year for date format is "fake" placeholder
  ggplot(aes(x = as.Date(paste(2014,strftime(date,format="%m-%d"),sep="-")),
                           y = gm30, 
                           color = as.factor(year))) +
  geom_jitter(size = 2) +
  facet_grid(. ~ location, scales = "free_y", labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("30-Day Geometric Mean Sample\nConcentration (CFU/100 mL)") +
  ylim(0,250) +
  scale_color_discrete(name = "Year") +

  # 30 day geometric mean threshold
  geom_hline(aes(yintercept = 35, linetype =  "Raw seafood consumption\n(35 CFU/100 mL)"), color = "blue") +
  scale_linetype_manual(name = "Threshold", values = c(2, 2),
                        guide = guide_legend(override.aes = list(color = c("blue")))) +
  # overall theme
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1,hjust = 1, angle = 35),
        strip.text.y = element_text(angle = 0)) +
  ggtitle("Enterococci\n30-Day Geometric Mean Concentrations") +
  points_theme

# save plot
ggsave("output/geomean_enterococci_conc_2019_2020.png", width = 10.5, height = 4)



# plot, fecal coliform 30 day geometric means
dat_excd %>%
  filter(data_type == "Fecal Coliform") %>%
  # note; use of year for date format is "fake" placeholder
  ggplot(aes(x = as.Date(paste(2014,strftime(date,format="%m-%d"),sep="-")),
                           y = gm30, 
                           color = as.factor(year))) +
  geom_jitter(size = 2) +
  facet_grid(. ~ location, scales = "free_y", labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("30-Day Geometric Mean Sample\nConcentration (CFU/100 mL)") +
  scale_color_discrete(name = "Year") +

  # 30 day geometric mean thresholds
  geom_hline(aes(yintercept = 14, linetype =  "Raw Seafood Consumption\n(14 CFU/100 mL)"), color = "blue") +
  geom_hline(aes(yintercept = 200, linetype =  "Water recreation, secondary\n(200 CFU/100 mL)"), color = "red") +
  scale_linetype_manual(name = "Threshold", values = c(2, 2),
                        guide = guide_legend(override.aes = list(color = c("blue","red")))) +
  # overall theme
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1,hjust = 1, angle = 35),
        strip.text.y = element_text(angle = 0)) +
  points_theme +
  ggtitle("Fecal Coliform\n30-Day Geometric Mean of Concentrations")

# save plot
ggsave("output/geomean_fecal_coliform_conc_2019_2020.png", width = 10, height = 4)



```



<br>



##### Table, frequency of 30-day rolling geometric mean exceedences

<br>

##### Figure, frequency of 30-day rolling geometric mean exceedences

Calculate % exceedence frequency by site, year, and bacteria type for 30-day rolling geometric mean

```{r echo = F, eval= FALSE}

### code examples

# plot
mst_tbl %>%
  ggplot(aes(as.factor(date),val, fill = analytical_method)) +
  geom_col(position = "dodge") +
  facet_grid(.~location, labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("Concentration (Copies/100 mL)") +
  theme_bw() +
   # modify legend
  scale_fill_discrete(name="Source",
                         breaks=c("Dog_BacCan-UCD", "Gull_Gull-4", "Human_HF183"),
                         labels=c("Dog", "Gull", "Human")) +
  # theme 
  theme(axis.text.x = element_text(vjust = .5, angle = 85),
        strip.text.y = element_text(angle = 0)) +
  col_theme



#### working here 12/10/20


# enterococci
## single-sample threshold
  geom_hline(aes(yintercept = 130, linetype =  "Raw seafood consumption\n(130 CFU/100 mL)"), color =


### FC
# single-sample thresholds
  geom_hline(aes(yintercept = 31, linetype =  "Raw Seafood Consumption\n(31 CFU/100 mL)"), color = "blue") +
  geom_hline(aes(yintercept = 400, linetype =  "Water recreation, secondary\n(400 CFU/100 mL)"), color = "red") +

```

need to make text larger on all figures plot of 24 hr sampling event in 2019??

<br>

### Other Exploratory Analysis

Is there a relationship between fecal coliform and enterococci concentrations?

```{r}


```
