---
title: "KWF_Beach_Sampling_Analysis_2019_2020"
output:
  html_document: 
    fig_width: 10
    fig_height: 6
    fig_caption: yes
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scroll: no
editor_options: 
  chunk_output_type: inline
---

Document last updated `r Sys.time()` by Benjamin Meyer ([ben\@kenaiwatershed.org](mailto:ben@kenaiwatershed.org){.email})

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

# clear environment
rm(list=ls())

# load packages
library(tidyverse)
library(lubridate)
library(readr)
library(readxl)
library(writexl)
library(hms)
library(plotly)
library(DT)
library(xlsx)
library(leaflet)
library(DT)
library(ggpubr)
library(plotrix)
library(zoo)
library(psych)
library(janitor)

select <- dplyr::select

# set plotting themes

## geom_col plots theme
col_theme <- theme(axis.title = element_text(size = 14, face = "bold"),
                   strip.text = element_text(size = 14, face = "bold"),
                   legend.title = element_text(size = 14, face = "bold"),
                   legend.text = element_text(size = 14),
                   axis.text = element_text(size = 14))

## geom_points plots theme
points_theme <- theme(axis.title = element_text(size = 14, face = "bold"),
                   strip.text = element_text(size = 14, face = "bold"),
                   legend.title = element_text(size = 14, face = "bold"),
                   legend.text = element_text(size = 14),
                   axis.text = element_text(size = 14),
                   title = element_text(size = 18))
```

<br>

This draft document contains preliminary data explorations of 2019-2020 beach bacteria sampling data from the Kenai River Watershed.

<br>

Notes:

-   Attempted to source beach sampling data from EPA water quality portal but did not find it there. Sourced data instead from AWQMS files and other files found on KWF server.

-   2020 MST data is present in AWQMS files, 2019 MST data is not. Sourced 2019 MST data from "MST results 2019.xlsx"

-   In the future, request that SGS can provide sample results in .csv or .xlsx format in addition to the PDFs to eliminate potential for data entry or data management error.

<br>

------------------------------------------------------------------------

### Read in data

```{r}

# import known data from local server source

# specify column import names
vars <- as.character(c("Monitoring Location Name",
            "Activity Latitude",
            "Activity Longitude",
            "Activity Medium",
            "Activity Start Date",
            "Activity Start Time",
            "Characteristic Name",
            "Result Analytical Method",
            "Result Value",
            "Result Value Units",
            "Result Qualifier"))

# specify new column names
new_vars <- as.character(c("location",
            "lat",
            "long",
            "medium",
            "date",
            "time",
            "data_type",
            "analytical_method",
            "val",
            "units",
            "qualifier"))


# import 2019 AWQMS data
## used this excel file because it had the most recent modified date
## see notes in excel file; not sure if all available data is contained within here
bs19 <- read_excel("data/KWF_2019_Kenai-BEACH_AWQMS_forQAQC_MH101119.xlsx", sheet = "AWQMS Data Template", skip = 7) %>%
  select(all_of(vars))

# import 2020 AWQMS data
## see notes in excel file; not sure if all avilable data is contained within here
bs20 <- read_excel("data/2020_KWF_Kenai-BEACH_AWQMS_SUBMITTED.xlsx", sheet = "RAW AWQMS Data Template", skip = 3) %>%
  select(all_of(vars))

# combine both years
dat <- bind_rows(bs19,bs20)

# rename columns
colnames(dat) <- new_vars

## 2019 MST data absent from AWQMS file.  So, import separately and format to join
## imported sheet was created manually on 12/8/20 from the sheet titled "Summary"
mst19 <- read_excel("data/MST results 2019.xlsx", sheet = "AWQMS_summary_format") %>%
  select(-site_id)

# join 2019 MST data to overall data set
dat <- bind_rows(dat,mst19)

# tidy column formats
dat <- dat %>%
  transform(time = hms::as_hms(time))

# replace "Gull_Gull-#" with "Gull_Gull-4"
# correct location names
dat <- dat %>%
  mutate(analytical_method = str_replace_all(analytical_method,c("Gull_Gull-5" = "Gull_Gull-4",
                                                                 "Gull_Gull-6" = "Gull_Gull-4"))) %>%
  mutate(location = str_replace_all(location, c("North Kenai River Beach 4" = "North Kenai Beach 4"))) %>%
  # include only water quality data
  # bacteria concs and temperature
  filter(medium == "Water")

rm(bs19,bs20,mst19)
                                             
```

<br>

Notes post data read-in

-   Air temperature values are all negative integers (maybe a code? do not use air temperature data from this data set)
-   Applied name correction to reconcile equivalent "North Kenai Beach 4" and "North Kenai River Beach 4" sites.
-   Some unit designations are missing (turbidity, weather condition, wind direction, others?)
-   Some Microbial Source Tracking analytic method ID is entered as "Gull_Gull-\#" (e.g. 5, 6, 7, etc) when the only designation for this method should be "Gull_Gull-4." Likely occurred from an excel drag-down at some point in data entry.
-   Many observations in "val" column are character (e.g. "MODERATE", "ND", etc). Caution when converting to numeric.
-   Verified that "NA" observations in the value column are actually absent, checked original PDF files from SGS.
-   Replicate values are present

<br>

Procedure:

-   Recreate tables (approximately) found in provided excel file ("2020 Beach Sampling Compiled Results_FINALREPORT.xlsx") to verify values
-   Create figures to visualize exceedence data

<br>

------------------------------------------------------------------------

<br>

### Replicate sample values

Address replicate sample values

-   How many replicate samples do we have?

```{r}
# how many replicate samples do we have?
rep_ct <- dat %>%
  filter(data_type %in% c("Enterococci","Fecal Coliform","Microbial Source Tracking")) %>%
  group_by(location,date,time,data_type) %>%
  count(name = "rep_set") %>%
  ungroup() %>%
  filter(rep_set > 1) %>%
  group_by(data_type,rep_set) %>%
  count(name = "count")

datatable(rep_ct)

  
```

The 2019 - 2020 data set has a number of sample values where a paired replicate also exists, both for bacteria concentration values and microbial source tracking values. 

As per ADEC CALM protocol, where two repicate values exist, only the higher of the two values are used in further analyses.

<br>

```{r}
# mst sample values are in sets of 3 and have alphabetical characters for concentrations, so it treat separately
mst_dat <- dat %>%
  filter(data_type == "Microbial Source Tracking")


# prep bacteria concentrations dataframe
dat  <- dat %>%
  filter(data_type %in% c("Enterococci","Fecal Coliform")) %>%  
  group_by(location,lat,long,medium,data_type,analytical_method,units,qualifier,date,time) %>%
  
  # where pairs of two replicate samples exist, retain only the higher of the two values.  
  ## For example: South Kenai Beach 3 Enterococci 2019-05-21, rep1 = 4, rep 2 = 5; retains only rep1 = 4
  filter(val == max(val)) %>%
  
  # in some cases, when two replicate values are present, they are identical (e.g. rep1 = 1, rep2 = 1).  where equivalencies exist, we want to retain only one of the values
  ungroup() %>%
  group_by(location,lat,long,medium,data_type,analytical_method,units,qualifier,date,time,val) %>%
  summarise(val = mean(as.numeric(val))) %>%
  
  # and, in some cases, where two identical values exist, one may have a qualifier (e.g. "U") and the other does not
  ## We want to retain the one without the qualifier (e.g. "higher" value)
  ## remove 'qualifier' as grouping factor but keep it as a column
  ungroup() %>%
  distinct(location,lat,long,medium,data_type,analytical_method,units,date,time,val, .keep_all = T)

```

### Microbial Source Tracking

#### MST Data Table

```{r}
# recreate table in tab "2019-2020_FinalReport_MST2020"
mst_tbl <- mst_dat %>%
  arrange(location,date,analytical_method) %>%
  select(date,location,analytical_method,val) %>%
  pivot_wider(names_from = analytical_method, values_from = val) %>%
  arrange(date,location)

colnames(mst_tbl) <- c("Date","Location","Dog Feces","Gull Feces","Human Feces")

# print
datatable(mst_tbl,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))

```

<br>

#### MST Results Figure

```{r}
# prep table for plot
mst_tbl <- mst_dat %>%
  filter(data_type == "Microbial Source Tracking") %>%
  arrange(location,date,analytical_method) %>%
  select(date,location,analytical_method,val) %>%
  # replace results DNQ and ND with zero and convert to numeric
  mutate(val = as.numeric(str_replace_all(val,c("DNQ" = "0", "ND" = "0"))))

# plot
mst_tbl %>%
  ggplot(aes(as.factor(date),val, fill = analytical_method)) +
  geom_col(position = "dodge") +
  facet_grid(.~location, labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("Concentration (Copies/100 mL)") +
  theme_bw() +
   # modify legend
  scale_fill_discrete(name="Source",
                         breaks=c("Dog_BacCan-UCD", "Gull_Gull-4", "Human_HF183"),
                         labels=c("Dog", "Gull", "Human")) +
  # theme 
  theme(axis.text.x = element_text(vjust = .5, angle = 85),
        strip.text.y = element_text(angle = 0)) +
  col_theme

# save plot
ggsave("output/mst_kenai_beach_2019_2020.png", width = 12, height = 6)
```

<br>

------------------------------------------------------------------------

### Enterococci and Fecal Coliform Concentrations

#### Bacteria Concentrations Data Table

```{r}
# prep table
bac_tbl <- dat %>% 
  filter(data_type %in% c("Enterococci","Fecal Coliform")) %>% 
  arrange(location,date,data_type) %>%
  select(date,location,data_type,val) %>%
  # replace results DNQ and ND with zero and convert to numeric
  mutate(val = as.numeric(str_replace_all(val,c("DNQ" = "0", "ND" = "0")))) %>%
  # use day of year for x axis and year for facet grid
  mutate(day = yday(date),
         year = year(date))

# print
datatable(bac_tbl,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))

```

<br>

#### Bacteria Concentrations Figures

```{r, fig.height=10, fig.width=12}
# plot 1
bac_tbl %>%
  ggplot(aes(x = as.Date(paste(2014,strftime(date,format="%m-%d"),sep="-")),
                           y = val, 
                           color = as.factor(year))) +
  geom_jitter(size = 3) +
  facet_grid(location~data_type, labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("Individual Sample Concentration (CFU/100 mL)") +
  scale_color_discrete(name = "Year")+
  # theme
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1),
        strip.text.y = element_text(angle = 0)) +
  points_theme +
  ggtitle("Bacteria Raw Concentrations")

# save plot
ggsave("output/bacteria_conc_2019_2020_v1.png", width = 10, height = 10)

```

```{r}
# plot 2, enterococci
bac_tbl %>%
  filter(data_type == "Enterococci") %>%
  # note; use of year for date format is "fake" placeholder
  ggplot(aes(x = as.Date(paste(2014,strftime(date,format="%m-%d"),sep="-")),
                           y = val, 
                           color = as.factor(year))) +
  geom_jitter(size = 2) +
  facet_grid(. ~ location, scales = "free_y", labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("Individual Sample\nConcentration (CFU/100 mL)") +
  ylim(0,625) +
  ylim(0,625) +
  scale_color_discrete(name = "Year") +

  # single-sample threshold
  geom_hline(aes(yintercept = 130, linetype =  "Raw seafood consumption\n(130 CFU/100 mL)"), color = "blue") +
  scale_linetype_manual(name = "Threshold", values = c(2, 2),
                        guide = guide_legend(override.aes = list(color = c("blue")))) +
  # overall theme
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1,hjust = 1, angle = 35),
        strip.text.y = element_text(angle = 0)) +
  ggtitle("Enterococci Concentrations") +
  points_theme

# save plot
ggsave("output/enterococci_conc_2019_2020.png", width = 10, height = 4)


# plot 3, fecal coliform
bac_tbl %>%
  filter(data_type == "Fecal Coliform") %>%
  # note; use of year for date format is "fake" placeholder
  ggplot(aes(x = as.Date(paste(2014,strftime(date,format="%m-%d"),sep="-")),
                           y = val, 
                           color = as.factor(year))) +
  geom_jitter(size = 2) +
  facet_grid(. ~ location, scales = "free_y", labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("Individual Sample\nConcentration (CFU/100 mL)") +
  scale_color_discrete(name = "Year") +

  # single-sample thresholds
  geom_hline(aes(yintercept = 31, linetype =  "Raw Seafood Consumption\n(31 CFU/100 mL)"), color = "blue") +
  geom_hline(aes(yintercept = 400, linetype =  "Water recreation, secondary\n(400 CFU/100 mL)"), color = "red") +
  scale_linetype_manual(name = "Threshold", values = c(2, 2),
                        guide = guide_legend(override.aes = list(color = c("blue","red")))) +
  # overall theme
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1,hjust = 1, angle = 35),
        strip.text.y = element_text(angle = 0)) +
  points_theme +
  ggtitle("Fecal Coliform Concentrations")

# save plot
ggsave("output/fecal_coliform_conc_2019_2020.png", width = 10, height = 4)

```

<br>

------------------------------------------------------------------------

### Exeedence Frequency

Define ADEC threshold values

-   Enterococci:

    -   "In a 30-day period, the geometric mean of samples may not exceed 35 enterococci CFU/100 mL, and not more than 10% of the samples may exceed a statistical threshold value (STV) of 130 CFU/100 mL."

-   Fecal coliform:

    -   "Harvest for consumption of raw mollusks or other raw seafood: The geometric mean of samples may not exceed 14 fecal coliform/100 ml; and not more than 10% of the samples may exceed 31 CFU/100 mL for a membrane filtration test."

    -   "Water recreation, secondary: In a 30-day period, the geometric mean of samples may not exceed 200 fecal coliform/100 mL, and not more than 10% of the samples may exceed 400 fecal coliform/100 mL."

<br>

General Data Notes

-   For geometric mean calculations, non-detect values were assigned 50% their method detection limit: See section 2.4.1 of ADEC "Listing Methodology for Determining Water Quality Impairments from Pathogens": "It is appropriate to consider non-detect samples when calculating the geometric mean for assessment purposes. Rather than eliminating the "non-detects" from the assessment data, these results and sample results measured below the detection limit will be calculated as 50% of the method detection limit. This approach may not be appropriate during the analysis of water quality trends."

<br>

Calculate 30-day rolling geometric mean values for Enterococci and Fecal Coliform concentrations

```{r}
dat  <- dat %>%
  filter(data_type %in% c("Fecal Coliform","Enterococci")) %>%
  mutate(year = year(date)) %>%
  transform(val = as.numeric(val)) %>%
  # use 50% of value if it has an "Undetected" qualifier
  mutate(val = ifelse(is.na(qualifier),val,
                      ifelse(qualifier == "U",(val*.5),val))) %>%
  group_by(data_type,location,year) %>%
  arrange(year,data_type,location,date) %>%
  # calculate 30 day rolling geometric means
  mutate(gm30 = rollapplyr(val, 1:n() - findInterval(date - days(30), date), 
                           geometric.mean, fill = NA, partial = T)) %>%
  
  # mutate values from the first annual 30 days of sampling to NA. (We need 30 days for a 30-day rolling geometric mean value).
  mutate(gm30 = as.numeric(ifelse(date < min(date + days(30)),"",gm30)))

```

<br>

```{r}
# create columns of exceedence occurences
dat  <- dat %>%
  # individual sample exceedences
  mutate(indv_excd = case_when(
    (val < 130 & data_type == "Enterococci") ~ "no_indv_excd",
    (val >= 130 & data_type == "Enterococci") ~ "ec_excd_130",
    (val < 31 & data_type == "Fecal Coliform") ~ "no_indv_excd",
    (val >= 31 & val < 400 & data_type == "Fecal Coliform") ~ "fc_excd_31",
    (val >= 400 & data_type == "Fecal Coliform") ~ "fc_excd_400")) %>%
  # 30-day rolling geometric mean value exceedences
  mutate(gm30_excd = case_when(
    (gm30 < 35 & data_type == "Enterococci") ~ "no_gm_excd",
    (gm30 >= 35 & data_type == "Enterococci") ~ "gm_ec_excd_35",
    (gm30 < 14 & data_type == "Fecal Coliform") ~ "no_gm_excd",
    (gm30 >= 14 & val < 200 & data_type == "Fecal Coliform") ~ "gm_fc_excd_14",
    (gm30 >= 200 & data_type == "Fecal Coliform") ~ "gm_fc_excd_200"))
```

<br>

Calculate percent of individual sample values in rolling 30-day period exceeding thresholds

```{r}

z <- dat %>%
  filter(!is.na(val)) %>%
  group_by(data_type,location,year) %>%
  mutate(obs = 1) %>%

  # mutate if individual exceedence present
  mutate(indv_ec_excd_130 = ifelse(indv_excd == "ec_excd_130",1,0),
         indv_fc_excd_31  = ifelse(indv_excd == "fc_excd_31",1,0),
         indv_fc_excd_400 = ifelse(indv_excd == "fc_excd_400",1,0)) %>%
  
  # mutate if gm30 exceedence present
  mutate(gm30_ec_excd_35 = ifelse(gm30_excd == "gm_ec_excd_35",1,0),
         gm30_fc_excd_14  = ifelse(gm30_excd == "gm_fc_excd_14",1,0),
         gm30_fc_excd_200 = ifelse(gm30_excd == "gm_fc_excd_200",1,0)) %>%
  
  
  ungroup() %>%
  group_by(data_type,location,year,gm30,
           indv_ec_excd_130,indv_fc_excd_31,indv_fc_excd_400,
           gm30_ec_excd_35, gm30_fc_excd_14, gm30_fc_excd_200) %>%
  
 # count samples from each date (some duplicates)
  count(date, wt = obs, name = "obs") %>%
  arrange(data_type,location,year,date) %>%
  
 # sum the number of observations available from previous 30 days
 # these 3 lines of code took me 5 hours to figure out :/
  ungroup() %>%
  group_by(data_type,location,year) %>%
  mutate(grp_sample_ct = rollapplyr(obs, 1:n() - findInterval(date - days(30), date), 
                           sum, fill = NA, partial = T)) %>%
  
  # retain only instances with 30 days of prior observations
  filter(!is.na(gm30)) %>%
  
  # calculate percentage of individual samples exceeding various thresholds in previous 30 days
  
  ## instances in which 10% of samples from past 30 days exceed thresholds
  mutate(pct_indv_ec_excd_130 = indv_ec_excd_130 / grp_sample_ct,
         pct_indv_fc_excd_31 = indv_fc_excd_31 / grp_sample_ct,
         pct_indv_fc_excd_400 = indv_fc_excd_400 / grp_sample_ct) %>%
  
  # retain desired columns
  select(data_type,location,year,date,contains("gm30"),contains("pct")) %>%
  ungroup() %>%
  arrange(data_type,location,year,date)


# want to retain gm30 column alongside others; currently stacked

## messy steps below

## discovered here:  there are some repeat days w/ diff gm30 values.  need to check earlier grouping factors; like the ones used to calc gm30.  will make possible to use spread fxn correctly once resolved



z1 <- z %>%
  select(data_type,date,location,gm30) %>%
  spread(data_type,gm30)
  
ec <- z %>%
  filter(data_type == "Enterococci") %>%
  select(location,date,gm30) %>%
  rename(gm30_ec = gm30)

fc <- z %>%
  filter(data_type == "Fecal Coliform") %>%
  select(location,date,gm30) %>%
  rename(gm30_fc = gm30)

z1 <- left_join(ec,fc,by = c("location","date"))


# final table to print; export to excel file? or can r generate a color scheme down-column like excel?
x <- left_join(z,z1)

# two gm30 cells are NA but other column shows an exceedence on that day... find out why
# final:  publish each site as an indv page, highlight cell values w/ exceedences; pair w/ horizon geom segment vs day of



  

```

<br>

Calculate % exceedence frequency by site, year, and bacteria type for individual samples

```{r}
  library(janitor)

# prop table for individual exceedences
indv_excd_tbl <- dat_excd %>%
  # remove observations with no value = NA
  filter(!is.na(val)) %>%
  select(location,data_type,year,indv_excd) %>%
  group_by(location,data_type,year) %>%
  count(indv_excd) %>%
  spread(key = indv_excd, val = n) %>%
  mutate_at(vars(contains("excd")),funs(as.numeric)) %>%
  mutate(across(everything(), ~replace_na(.x, 0))) %>%
  # sum across columns
  # was unable to find a way to successfully use a function to sum across columns here; used manual addition...
  mutate(sample_n = ec_excd_130 + fc_excd_31 + fc_excd_400 + no_indv_excd) %>%
  # calculate % of sample ct exceedence
  mutate(ec_excd_130 = ec_excd_130 / sample_n,
         fc_excd_31 = fc_excd_31 / sample_n,
         fc_excd_400 = fc_excd_400 / sample_n) %>%
  select(-no_indv_excd) 

```

<br>

#### Individual sample value exceedences

<br>

##### Table, overall frequency of individual sample exceedences

```{r}
indv_excd_tbl_exp <- indv_excd_tbl %>%
  select(location, data_type,year,sample_n,ec_excd_130,fc_excd_31,fc_excd_400) %>%
  transform(year = as.factor(year),
            sample_n = as.factor(sample_n)) %>%
  adorn_pct_formatting()

# prep table for export
colnames(indv_excd_tbl_exp) <- c("Location","Bacteria","Year","Sample Count",
                                 "% Enterococci Samples Exceeding 130 CFU/100 mL",
                                 "% Fecal Coliform Samples Exceeding 31 CFU/100 mL",
                                 "% Fecal Coliform Samples Exceeding 400 CFU/100 mL")
# save table in csv
write_csv(indv_excd_tbl_exp,"output/individual_sample_exceedences.csv")

# print table
datatable(indv_excd_tbl_exp,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))

```

<br>

##### Table, frequency of individual sample exceedences within 30-day periods

Recall; regarding individual sample values and 30-day periods:

Enterococci:

-   "... not more than 10% of the samples may exceed a statistical threshold value (STV) of 130 CFU/100 mL."

Fecal coliform:

-   "... not more than 10% of the samples may exceed 31 CFU/100 mL for a membrane filtration test."

-   "... not more than 10% of the samples may exceed 400 fecal coliform/100 mL."

```{r}
# calculate % of individual samples exceeding thresholds within 30-day periods
indv_excd_30day <- dat_excd %>%
  # designate 30-day periods for each site/year
  group_by(location,data_type,year) %>%
  mutate(period_30day = rep(seq_len(n()/2), each = 4, length.out = n()),
         sample_ct = n()) %>%
  ungroup() %>%
  # create columns of start/end dates for each 30-day period
  group_by(location,data_type,year,period_30day) %>%
  mutate(start_date = min(date),
         end_date = max(date)) %>%
  # count number of exceedences by type, site, year
  group_by(location,data_type,year,sample_ct,period_30day,start_date,end_date) %>%
  filter(!is.na(indv_excd)) %>%
  count(indv_excd) %>%
  # what % of individual samples exceeded thresholds during 30-day periods?
  mutate(pct_indv_samples_excd = (n / sample_ct)) %>%
  # remove instances of no individual sample exceedences
  filter(indv_excd != "no_indv_excd") 

# prep table for export
indv_excd_30day_exp <- indv_excd_30day %>%
  # retain only instances where > 10% of individual samples exceeded thresholds
  filter(pct_indv_samples_excd > .1)  %>%
  ungroup() %>%
  dplyr::mutate_all(as.character) %>%
  mutate(pct_indv_samples_excd = as.numeric(pct_indv_samples_excd)) %>%
  # adorn with percentage format
  adorn_pct_formatting() %>%
  select(location,data_type,year,indv_excd,start_date,end_date,pct_indv_samples_excd) %>%
  # replace exceedence type with longer names
  mutate(indv_excd = str_replace_all(indv_excd,c("fc_excd_31" = "Fecal Coliform > 31",
                                                                 "fc_excd_400" = "Fecal Coliform > 400",
                                                                 "ec_excd_130" = "Enterococci > 130"))) %>%
  spread(indv_excd,pct_indv_samples_excd) %>%
  arrange(year,location,data_type)

 # rename columns  
colnames(indv_excd_30day_exp) <- c("Location","Bacteria","Year","Start Date","End Date",
                                   "% Samples with Enterococci > 130 in 30-day period",
                                   "% Samples with Fecal Coliform > 31 in 30-day period")

# save table in csv
write.csv(indv_excd_30day_exp,"output/30day_periods_individual_sample_exceedences.csv", row.names = F)

# print table
datatable(indv_excd_30day_exp,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))

```

<br>

##### Figure, frequency of individual sample exceedences

```{r}
indv_excd_tbl %>%
  gather(excd, val, contains("excd")) %>%
  ggplot(aes(as.factor(year),val,fill = excd)) +
  geom_col(position = "dodge2") +
  facet_grid(. ~ location, labeller = labeller(location = label_wrap_gen(5))) +
  ylim(0,1) +
  xlab("") +
  ylab("% of Total Indvidual Samples\nExceeding Thresholds") + 
  theme_bw() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  # modify legend text
  scale_fill_discrete(name="Exceedence Type\n(CFU/100 mL)",
                      breaks = c("ec_excd_130","fc_excd_31","fc_excd_400"),
                      labels = c("Enterococci > 130",
                                 "Fecal Coliform > 31",
                                 "Fecal Coliform > 400")) +
  col_theme

# save image
ggsave("output/individual_sample_exceedences.png",width = 10, height = 4)

```

<br>

##### Figure, frequency of individual sample exceedences within unique 30-day periods

```{r}
indv_excd_30day %>%
  ggplot(aes(period_30day,pct_indv_samples_excd,fill = indv_excd)) +
  geom_col(position = position_dodge(preserve = "single")) +
  facet_grid(year ~ location, labeller = labeller(location = label_wrap_gen(5))) +
  ylim(0,.3) +
  xlab("30-Day Sampling Period") +
  ylab("% of Indvidual Samples\nExceeding Thresholds Within 30-Day Periods") + 
  theme_bw() +
  geom_hline(yintercept = .1, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  # modify legend text
  scale_fill_discrete(name="Exceedence Type\n(CFU/100 mL)",
                      breaks = c("ec_excd_130","fc_excd_31","fc_excd_400"),
                      labels = c("Enterococci > 130",
                                 "Fecal Coliform > 31",
                                 "Fecal Coliform > 400")) +
  col_theme

# save image
ggsave("output/30day_individual_sample_exceedences.png",width = 10, height = 4)


```

<br>

------------------------------------------------------------------------

------------------------------------------------------------------------

<br>

#### 30-day rolling geometric mean of sample values, exceedences

##### Figure, individual exceedences, 30-day rolling geometric mean

```{r}

# plot, enterococci 30 day geometric means
dat_excd %>%
  filter(data_type == "Enterococci") %>%
  # note; use of year for date format is "fake" placeholder
  ggplot(aes(x = as.Date(paste(2014,strftime(date,format="%m-%d"),sep="-")),
                           y = gm30, 
                           color = as.factor(year))) +
  geom_jitter(size = 2) +
  facet_grid(. ~ location, scales = "free_y", labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("30-Day Geometric Mean Sample\nConcentration (CFU/100 mL)") +
  ylim(0,250) +
  scale_color_discrete(name = "Year") +

  # 30 day geometric mean threshold
  geom_hline(aes(yintercept = 35, linetype =  "Raw seafood consumption\n(35 CFU/100 mL)"), color = "blue") +
  scale_linetype_manual(name = "Threshold", values = c(2, 2),
                        guide = guide_legend(override.aes = list(color = c("blue")))) +
  # overall theme
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1,hjust = 1, angle = 35),
        strip.text.y = element_text(angle = 0)) +
  ggtitle("Enterococci\n30-Day Geometric Mean Concentrations") +
  points_theme

# save plot
ggsave("output/geomean_enterococci_conc_2019_2020.png", width = 10.5, height = 4)



# plot, fecal coliform 30 day geometric means
dat_excd %>%
  filter(data_type == "Fecal Coliform") %>%
  # note; use of year for date format is "fake" placeholder
  ggplot(aes(x = as.Date(paste(2014,strftime(date,format="%m-%d"),sep="-")),
                           y = gm30, 
                           color = as.factor(year))) +
  geom_jitter(size = 2) +
  facet_grid(. ~ location, scales = "free_y", labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("30-Day Geometric Mean Sample\nConcentration (CFU/100 mL)") +
  scale_color_discrete(name = "Year") +

  # 30 day geometric mean thresholds
  geom_hline(aes(yintercept = 14, linetype =  "Raw Seafood Consumption\n(14 CFU/100 mL)"), color = "blue") +
  geom_hline(aes(yintercept = 200, linetype =  "Water recreation, secondary\n(200 CFU/100 mL)"), color = "red") +
  scale_linetype_manual(name = "Threshold", values = c(2, 2),
                        guide = guide_legend(override.aes = list(color = c("blue","red")))) +
  # overall theme
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1,hjust = 1, angle = 35),
        strip.text.y = element_text(angle = 0)) +
  points_theme +
  ggtitle("Fecal Coliform\n30-Day Geometric Mean of Concentrations")

# save plot
ggsave("output/geomean_fecal_coliform_conc_2019_2020.png", width = 10, height = 4)



```

<br>

##### Table, frequency of 30-day rolling geometric mean exceedences

Calculate % exceedence frequency by site, year, and bacteria type for 30-day rolling geometric mean values

```{r}

# prop table for individual exceedences
gm30_excd_tbl <- dat_excd %>%
  # remove observations with no value = NA
  filter(!is.na(val),
         !is.na(gm30_excd)) %>%
  select(location,data_type,year,gm30_excd) %>%
  group_by(location,data_type,year) %>%
  count(gm30_excd) %>%
  spread(key = gm30_excd, val = n) %>%
  mutate_at(vars(contains("excd")),funs(as.numeric)) %>%
  mutate(across(everything(), ~replace_na(.x, 0))) %>%
  # sum across columns
  # was unable to find a way to successfully use a function to sum across columns here; used manual addition...
  mutate(sample_n = gm_ec_excd_35 + gm_fc_excd_14 + gm_fc_excd_200 + no_gm_excd) %>%
  # calculate % of sample ct exceedence
  mutate(gm_ec_excd_35 = gm_ec_excd_35 / sample_n,
         gm_fc_excd_14 = gm_fc_excd_14 / sample_n,
         gm_fc_excd_200 = gm_fc_excd_200 / sample_n) %>%
  select(-no_gm_excd) 

# prep for display and export
gm30_excd_tbl_exp <- gm30_excd_tbl %>%
  select(location, data_type,year,sample_n,gm_ec_excd_35,gm_fc_excd_14,gm_fc_excd_200) %>%
  transform(year = as.factor(year),
            sample_n = as.factor(sample_n)) %>%
  adorn_pct_formatting()

# prep table for export
colnames(gm30_excd_tbl_exp) <- c("Location","Bacteria","Year","Sample Count",
                                 "% Enterococci 30-Day Geometric Means Exceeding 35 CFU/100 mL",
                                 "% Fecal Coliform 30-Day Geometric Means Exceeding 14 CFU/100 mL",
                                 "% Fecal Coliform 30-Day Geometric Means Exceeding 200 CFU/100 mL")


# save table in csv
write_csv(gm30_excd_tbl_exp,"output/gm30_value_exceedences.csv")

# print table
datatable(gm30_excd_tbl_exp,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))

```

<br>

##### Figure, frequency of 30-day rolling geometric mean value exceedences

```{r}

# calculate % of 30-day rolling geometric mean values exceeding thresholds by site/year/threshold type
gm30_excd <- dat_excd %>%
  # count number of exceedences by type, site, year
  group_by(location,data_type,year) %>%
  mutate(sample_ct = n()) %>%
  group_by(location,data_type,year,sample_ct) %>%
  filter(!is.na(gm30_excd)) %>%
  count(gm30_excd) %>%
  # what % of individual samples exceeded thresholds during 30-day periods?
  mutate(pct_gm30_val_excd = (n / sample_ct)) %>%
  # remove instances of no individual sample exceedences
  filter(gm30_excd != "no_gm_excd") 



# plot of % of 30-day rolling geomen values exceeding various thresholds
gm30_excd %>%
  ggplot(aes(as.factor(year),pct_gm30_val_excd,fill = gm30_excd)) +
  geom_col(position = position_dodge(preserve = "single")) +
  facet_grid(. ~ location, labeller = labeller(location = label_wrap_gen(5))) +
  #ylim(0,.3) +
  xlab("") +
  ylab("% of 30-Day Rolling Geometric Mean Values\nExceeding Thresholds") + 
  theme_bw() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  # modify legend text
  scale_fill_discrete(name="Exceedence Type\n(30-Day Rolling Geometric Mean\nCFU/100 mL)",
                      breaks = c("gm_ec_excd_35","gm_fc_excd_14","gm_fc_excd_200"),
                      labels = c("Enterococci > 35",
                                 "Fecal Coliform > 14",
                                 "Fecal Coliform > 200")) +
  col_theme

# save image
ggsave("output/gm30_val_exceedences.png",width = 10, height = 4)

```

Calculate % exceedence frequency by site, year, and bacteria type for 30-day rolling geometric mean

```{r echo = F, eval= FALSE}

### code examples

# plot
mst_tbl %>%
  ggplot(aes(as.factor(date),val, fill = analytical_method)) +
  geom_col(position = "dodge") +
  facet_grid(.~location, labeller = labeller(location = label_wrap_gen(5))) +
  xlab("") +
  ylab("Concentration (Copies/100 mL)") +
  theme_bw() +
   # modify legend
  scale_fill_discrete(name="Source",
                         breaks=c("Dog_BacCan-UCD", "Gull_Gull-4", "Human_HF183"),
                         labels=c("Dog", "Gull", "Human")) +
  # theme 
  theme(axis.text.x = element_text(vjust = .5, angle = 85),
        strip.text.y = element_text(angle = 0)) +
  col_theme



#### working here 12/10/20


# enterococci
## single-sample threshold
  geom_hline(aes(yintercept = 130, linetype =  "Raw seafood consumption\n(130 CFU/100 mL)"), color =


### FC
# single-sample thresholds
  geom_hline(aes(yintercept = 31, linetype =  "Raw Seafood Consumption\n(31 CFU/100 mL)"), color = "blue") +
  geom_hline(aes(yintercept = 400, linetype =  "Water recreation, secondary\n(400 CFU/100 mL)"), color = "red") +

```

need to make text larger on all figures plot of 24 hr sampling event in 2019??

<br>

### Other Exploratory Analysis

Is there a relationship between fecal coliform and enterococci concentrations?

```{r}


```
