---
title: "virtual_beach"
output:
  html_document:
    fig_width: 10
    fig_height: 6
    fig_caption: yes
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: inline
---

Document last updated `r Sys.time()` by Benjamin Meyer ([ben\@kenaiwatershed.org](mailto:ben@kenaiwatershed.org){.email})

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

# clear environment
rm(list=ls())

# load packages
library(tidyverse)
library(lubridate)
library(readr)
library(readxl)
library(writexl)
library(hms)
library(plotly)
library(DT)
library(xlsx)
library(leaflet)
library(DT)
library(ggpubr)
library(plotrix)
library(zoo)
library(psych)
library(janitor)
library(scales)
library(htmltools)

select <- dplyr::select

# set plotting themes

## geom_col plots theme
col_theme <- theme(axis.title = element_text(size = 14, face = "bold"),
                   strip.text = element_text(size = 14, face = "bold"),
                   legend.title = element_text(size = 14, face = "bold"),
                   legend.text = element_text(size = 14),
                   axis.text = element_text(size = 14))

## geom_points plots theme
points_theme <- theme(axis.title = element_text(size = 14, face = "bold"),
                   strip.text = element_text(size = 14, face = "bold"),
                   legend.title = element_text(size = 14, face = "bold"),
                   legend.text = element_text(size = 14),
                   axis.text = element_text(size = 14),
                   title = element_text(size = 18))

# function to exclude multiple items per column
'%ni%' <- Negate('%in%')
```

<br>

This draft document prepares and examines data from the EPA BEACON beach database that will be used in implementing the EPA Virtual Beach model for North and South Kenai Beach.

The script aims to produce site-specific spreadsheets of environmental data as described inthe Virtual Beach User manual (p 15,"Global Datasheet: Data Requirements and Considerations.")

<br>

General notes:

-   Environmental data was downloaded as a csv file from EPA BEACON repository using the "reports" function in December 2020 (<https://watersgeo.epa.gov/BEACON2/reports.html>). The file contains data all archived data 2010 - 2020.

-   This script is in development, using guidance from virtual beach user manual at <https://www.epa.gov/sites/production/files/2016-03/documents/vb3_manual_3.0.6.pdf>

-   World Meteorological Organization (WMO) 4501 weather codes at <https://www.jodc.go.jp/data_format/weather-code.html>

-   Watch out for mixed use of metric and standard units in original import from EPA database.

-   Need to get associated coordinates for each sampling site.

<br>

### Data import

Import and prepare csv file from EPA BEACON database in local project folder

```{r}
# choose variables to retain
vars <- as.character(c("Year",
                       "StationID.Def..",
                       "Station.Name..Def..",
                       "StartDate.Def..",
                       "StartTime.Def..",
                       "ActivityTypeCode.Def..",
                       "CharacteristicName.Def..",
                       "Result.Value.Def..",
                       "Result.MeasureUnit",
                       "Result.Comment.Def..",
                       "ActivityDepthValue.Def.."))

# rename variables
new_vars <- c("year",
              "station_id",
              "site",
              "date",
              "time",
              "sample_activity",
              "sample_type",
              "sample_value",
              "sample_unit",
              "sample_comment",
              "sample_depth_m")


# read in data
dat <- read.csv("data/epa_historical_data.csv") %>%
  select(all_of(vars))


# rename columns
colnames(dat) <- new_vars

# format col type 
dat <- dat %>%
  mutate(datetime = dmy_hms(paste(date,time))) %>%
  transform(date = date(datetime),
            time = hms::as_hms(datetime),
            sample_value = as.numeric(sample_value),
            sample_type = as.factor(sample_type)) %>%
  # remove missing values
  filter(!is.na(sample_value)) 

# convert all measurement units to METRIC for consistency... sheesh
dat <- dat %>%
  
  # farenheight to celsius
  mutate(sample_value = ifelse(sample_unit == "deg F", (sample_value - 32) / 1.8, sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "deg F", "deg C",sample_unit)) %>%
  
  # miles per hour to kph
  mutate(sample_value = ifelse(sample_unit == "mph",(sample_value*1.609344), sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "mph", "kph",sample_unit)) %>%
  
  # feet to meters
  mutate(sample_value = ifelse(sample_unit == "ft",(sample_value*0.3048), sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "ft", "m",sample_unit)) %>%
  
  # knots to meters per second
  mutate(sample_value = ifelse(sample_unit == "knots",(sample_value*0.51), sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "knots", "mps",sample_unit)) %>%
  
  # inches to meters
  
  
  # knots to meters per second
  mutate(sample_value = ifelse(sample_unit == "in",(sample_value*0.0254), sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "in", "m",sample_unit)) 



# some records of bacteria concentrations have inconsistent unit designations.  Each observation of enterococcus should be in "MPN/100 mL" and each observation of fecal coliform should be in "cfu/100 mL".

# apply bacteria unit corrections

dat <- dat %>%
  # fecal coliform units
  mutate(sample_unit = ifelse(sample_type == "Fecal Coliform","MPN/100ml",sample_unit)) %>%
  # enterococcus units
  mutate(sample_unit = ifelse(sample_type == "Enterococcus","cfu/100ml",sample_unit))


# modify and simplify sample type names
dat <- dat %>%
  mutate(sample_type = str_replace_all(
    sample_type,
    c("Temperature, air" = "air_temp_C",
      "Temperature, water" = "water_temp_C",
      "Fecal Coliform" = "fecal_coliform_cfu_100mL",
      "Enterococcus" = "enterococcus_mpn_100mL",
      "Tide range" = "tide_range_m",
      "Wind velocity" = "wind_velocity_kph",
      "Conductivity" = "conductivity_mS_cm",
      "Turbidity" = "turbidity_NTU",
      "Wave height" = "wave_height_m"
      )))

# some sample type names are stubborn and will not readily modify... may have hidden spaces or other formatting

# use another method

# filter and rename the stubborn sample types
weather <- dat %>%
  filter(sample_type == "Weather condition (WMO code 4501) (choice list)") %>%
  mutate(sample_type = "weather_code_wmo4501")

wind <- dat %>%
  filter(sample_type == "Wind direction (direction from, expressed 0-360 deg)") %>%
  mutate(sample_type = "wind_direction_deg")

# rejoin to overall dataset
dat <- dat %>%
  filter(sample_type %ni% c("Weather condition (WMO code 4501) (choice list)",
                            "Wind direction (direction from, expressed 0-360 deg)")) %>%
  bind_rows(weather,wind) %>%
  
  # remove missing dates
  filter(!is.na(date))


# clean up
rm(weather,wind)

```

<br>

### Data summary

What is the temporal extent for each type of data we have at each location?

```{r}
# summarise
tbl <- dat %>%
  # treat weather codes separately; they are not a continuous variable
  #filter(sample_type != "weather_code_wmo4501") %>%
  group_by(site,sample_type) %>%
  summarise(n = n(),
            mean = round(mean(sample_value),digits = 1),
            min = round(min(sample_value),digits = 1),
            max = round(max(sample_value),digits = 1),
            stderr = round(std.error(sample_value), digits = 1),
            min_year = min(year),
            max_year = max(year))

# print table
datatable(tbl,
          filter = 'top', 
          options = list(pageLength = 8, autoWidth = TRUE))

```

<br>

Notes:

-   There are some observations from 2010 at a site called "Kenai North Beach." This site name nomenclature is different than all the other years. Need to find out which "North Kenai Beach" site these data are from.

-   WMO Weather codes are not a continuous variable, so mean and std. error do not apply and are spurious values in the above table.

<br>

------------------------------------------------------------------------

#### Replicate samples

Examine presence of replicate samples in data set, and apply ADEC agency criteria for replicate sample values.

-   How many replicate samples do we have?

```{r}
# how many replicate samples do we have?
rep_ct <- dat %>%
  #filter(data_type %in% c("Enterococci","Fecal Coliform","Microbial Source Tracking")) %>%
  group_by(site,date,time,sample_type) %>%
  count(name = "rep_set") %>%
  ungroup() %>%
  filter(rep_set > 1) %>%
  group_by(sample_type,rep_set) %>%
  count(name = "count")

datatable(rep_ct)

  
```

Note on above table summarizing replicate samples:

-   For each sample type, "rep_set" indicates the quantity of simultaneously collected samples in a given instance, and "count" indicates the quantity of instances with simultaneous sample results. For example, as of Jan 15, 2021, we have 80 instances where 2 replicates of "enterococcus_mpn_100mL" exist.

-   To employ this data in the Virtual Beach model, we need a single value for each sample type and sampling event.

    -   For environmental variables (temperature, wind, etc), we will create an average value for each.

    -   For bacteria concentrations however, variation between replicates can be substantial.

-   ADEC protocol to address sample variability, as of Feb 2021 (personal communication, G. Augat, ADEC Juneau Office) is to only employ the "sample" value and not the "replicate" value.

-   In this data set we have additional instances where not just two replicate samples for bacteria concentrations exist, but also instances of 3, 4, 5, and 6 simultaneous replicates.

<br>

Apply criteria described above. Filter out samples that include the designation of either "Quality Control Sample-Blind Duplicate" or "Quality Control Field Replicate Msr/Obs"

```{r}
# create list of environmental variables
enviro_vars <- c("wind_velocity_kph",
                 "wave_height_m",
                 "water_temp_C",
                 "tide_range_m",
                 "air_temp_C",
                 "conductivity_mS_cm",
                 "pH",
                 "wind_direction_deg")

# max number of replicate environmental measurements is 2

# create means for environmental variables
enviro_dat <- dat %>%
  filter(sample_type %in% enviro_vars) %>%
  group_by(across(c(-sample_value,-sample_activity,-sample_comment,-sample_unit))) %>%
  summarise(mean_sample_value = mean(sample_value)) %>%
  ungroup()

# exclude samples designated as "Quality Control Sample-Blind Duplicate"

# and, if replicates still remain after that exclusion, retain only the highest value in instances where replicate bacteria concentrations values still exist

# rationale: some historical replicate samples may not have been recorded as such
bac_dat <- dat %>%
  filter(sample_type %in% c("fecal_coliform_cfu_100mL","enterococcus_mpn_100mL"),
         sample_activity != "Quality Control Sample-Blind Duplicate",
         sample_activity != "Quality Control Field Replicate Msr/Obs") %>%
  
  # group
  group_by(across(c(-sample_value,-sample_activity,-sample_comment))) %>%
  
  # where pairs of two replicate samples exist, retain only the higher of the two values.  
  filter(sample_value == max(sample_value)) %>%
  
  # if two identical replicate values are present, we want to retain only one of the values
  ungroup() %>%
  group_by(across(c(-sample_value,-sample_activity,-sample_comment))) %>%
  summarise(mean_sample_value = mean(sample_value))  

# rejoin bacteria and enviro data
dat <- bind_rows(enviro_dat,bac_dat)

# replace "NA" with blanks
z <- dat %>%
  replace(is.na(.), " ")
  

```

<br>

------------------------------------------------------------------------

### Data export

Create site-specific spreadsheets fo all data 2010 - 2020 following format described in Virtual Beach user guide (pg 15).

```{r }
# make all variables "wide" format so they each have their own column
dat <- dat %>%

  # remove unneeded columns for clean pivot
  ungroup() %>%
  select(site,date,time,datetime,sample_type,mean_sample_value) %>%
  distinct() %>%
  
  # make wider
  pivot_wider(names_from = sample_type,
              values_from = mean_sample_value) %>%


  # arrange rows chronologically by site
  arrange(site,datetime) 

# print table
datatable(dat,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))
  
```

<br>

------------------------------------------------------------------------

#### Site name consolidation

Explanation:

-   Original site names in the EPA database may be redundant for using the EPA virtual beach program For example, there are multiple "North Kenai Beach" and "South Kenai Beach" sites. The small difference is in coordinates for each of these sites are likely not relevant in this modeling application

<br>

Consolidate site names
```{r}
# apply site name consolidations
## read in manually created consolditaions
consolidate_sites <- read_excel("output/excel_tables/vb_site_names.xlsx")

## apply new site names

dat1 <- dat %>%
  left_join(consolidate_sites) %>%
  select(-site) %>%
  rename("site" = "consolidated_sitename") %>%
  relocate(site, .before = date) %>%
  # arrange chronologically by site
  arrange(site,datetime) %>%
  # create row number sets for each site
  group_by(site) %>%
  select(-date,-time) %>%
  relocate(datetime, .before = site)

```

<br>

#### Export to spreadsheets

Create separate csv spreadsheet exports for fecal coliform and enterococci and their associated environmental data

<br>

Enterococci export
```{r}
## remove old directory if present
unlink("output/virtual_beach_data/enterococci", recursive = T)
dir.create("output/virtual_beach_data/enterococci")

 
# write an individual csv for each site
dat1 %>%
  select(-fecal_coliform_cfu_100mL) %>%
  filter(!is.na(enterococcus_mpn_100mL)) %>%
  relocate(enterococcus_mpn_100mL, .after = datetime) %>%
  group_by(site) %>%
  do(write_csv(., paste0("output/virtual_beach_data/enterococci/",unique(.$site), ".csv")))

```
<br>

Fecal coliform export
```{r}
## remove old directory if present
unlink("output/virtual_beach_data/fecal_coliform", recursive = T)
dir.create("output/virtual_beach_data/fecal_coliform")

 
# write an individual csv for each site
dat1 %>%
  select(-enterococcus_mpn_100mL) %>%
  filter(!is.na(fecal_coliform_cfu_100mL)) %>%
  relocate(fecal_coliform_cfu_100mL, .after = datetime) %>%
  group_by(site) %>%
  do(write_csv(., paste0("output/virtual_beach_data/fecal_coliform/",unique(.$site), ".csv")))

```
