---
title: "virtual_beach"
output:
  html_document: 
    fig_width: 10
    fig_height: 6
    fig_caption: yes
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scroll: no
editor_options: 
  chunk_output_type: inline
---

Document last updated `r Sys.time()` by Benjamin Meyer ([ben\@kenaiwatershed.org](mailto:ben@kenaiwatershed.org){.email})

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

# clear environment
rm(list=ls())

# load packages
library(tidyverse)
library(lubridate)
library(readr)
library(readxl)
library(writexl)
library(hms)
library(plotly)
library(DT)
library(xlsx)
library(leaflet)
library(DT)
library(ggpubr)
library(plotrix)
library(zoo)
library(psych)
library(janitor)
library(scales)
library(naniar)

select <- dplyr::select

# set plotting themes

## geom_col plots theme
col_theme <- theme(axis.title = element_text(size = 14, face = "bold"),
                   strip.text = element_text(size = 14, face = "bold"),
                   legend.title = element_text(size = 14, face = "bold"),
                   legend.text = element_text(size = 14),
                   axis.text = element_text(size = 14))

## geom_points plots theme
points_theme <- theme(axis.title = element_text(size = 14, face = "bold"),
                   strip.text = element_text(size = 14, face = "bold"),
                   legend.title = element_text(size = 14, face = "bold"),
                   legend.text = element_text(size = 14),
                   axis.text = element_text(size = 14),
                   title = element_text(size = 18))

# function to exclude multiple items per column
'%ni%' <- Negate('%in%')
```

<br>

This draft document prepares and examines data from the EPA BEACON beach database that will be used in implementing the EPA Virtual Beach model for North and South Kenai Beach

<br>

Notes:

-   Data was downloaded as a csv file from EPA BEACON repository using the "reports" function in December 2020 (<https://watersgeo.epa.gov/BEACON2/reports.html>). Contains data 2012 - 2019.

-   To do: update data import process to include results from other years when available (pre-2012, and from 2020).

-   Script in development using guidance from virtual beach user manual at <https://www.epa.gov/sites/production/files/2016-03/documents/vb3_manual_3.0.6.pdf>

-   World Meteorological Organization (WMO) 4501 weather codes at https://www.jodc.go.jp/data_format/weather-code.html 

-   Watch out for mixed use of metric and standard units in original import from EPA database...

-   Need to get associated coordinates for each site

-   Data for Kenai Gull Rookery 2 and Warren Ames bridge sites not in output folder yet... where are they?  Maybe not in this DB on purpose?  Check w/ beacon db again.

<br>

### Data import

Import and prepare csv file from EPA BEACON database in local project folder

```{r}
# choose variables to retain
vars <- as.character(c("Year",
                       "StationID.Def..",
                       "Station.Name..Def..",
                       "StartDate.Def..",
                       "StartTime.Def..",
                       "ActivityTypeCode.Def..",
                       "CharacteristicName.Def..",
                       "Result.Value.Def..",
                       "Result.MeasureUnit",
                       "Result.Comment.Def..",
                       "ActivityDepthValue.Def.."))

# rename variables
new_vars <- c("year",
              "station_id",
              "site",
              "date",
              "time",
              "sample_activity",
              "sample_type",
              "sample_value",
              "sample_unit",
              "sample_comment",
              "sample_depth_m")


# read in data
dat <- read.csv("data/epa_historical_data.csv") %>%
  select(all_of(vars))

# rename columns
colnames(dat) <- new_vars

# format col type 
dat <- dat %>%
  mutate(datetime = dmy_hms(paste(date,time))) %>%
  transform(date = dmy(date),
            time = hms::as_hms(time),
            sample_value = as.numeric(sample_value),
            sample_type = as.factor(sample_type)) %>%
  # remove missing values
  filter(!is.na(sample_value)) 

# convert all measurement units to METRIC for consistency... sheesh
dat <- dat %>%
  
  # farenheight to celsius
  mutate(sample_value = ifelse(sample_unit == "deg F", (sample_value - 32) / 1.8, sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "deg F", "deg C",sample_unit)) %>%
  
  # miles per hour to kph
  mutate(sample_value = ifelse(sample_unit == "mph",(sample_value*1.609344), sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "mph", "kph",sample_unit)) %>%
  
  # feet to meters
  mutate(sample_value = ifelse(sample_unit == "ft",(sample_value*0.3048), sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "ft", "m",sample_unit)) %>%
  
  # knots to meters per second
  mutate(sample_value = ifelse(sample_unit == "knots",(sample_value*0.51), sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "knots", "mps",sample_unit)) 


# modify sample type names
dat <- dat %>%
  mutate(sample_type = str_replace_all(
    sample_type,
    c("Temperature, air" = "air_temp_C",
      "Temperature, water" = "water_temp_C",
      "Fecal Coliform" = "fecal_coliform_cfu_100mL",
      "Enterococcus" = "enterococcus_mpn_100mL",
      "Tide range" = "tide_range_m",
      "Wind velocity" = "wind_velocity_kph",
      "Conductivity" = "conductivity_mS_cm",
      "Turbidity" = "turbidity_NTU"
      )))

# some sample type names are stubborn and will not readily modify... may have hidden spaces or other formatting
# use another method
# filter and rename the stubborn sample types
weather <- dat %>%
  filter(sample_type == "Weather condition (WMO code 4501) (choice list)") %>%
  mutate(sample_type = "weather_code_wmo4501")

wind <- dat %>%
  filter(sample_type == "Wind direction (direction from, expressed 0-360 deg)") %>%
  mutate(sample_type = "wind_direction_deg")

# rejoin to overall dataset
dat <- dat %>%
  filter(sample_type %ni% c("Weather condition (WMO code 4501) (choice list)",
                            "Wind direction (direction from, expressed 0-360 deg)")) %>%
  bind_rows(weather,wind)

# make environmental variables "wide" format so they each have their own column
dat <- dat %>%
  mutate(row_id = row_number()) %>%
  # remove unneeded columns for clean pivot
  select(site,date,time,datetime,sample_type,sample_value) %>%
  
  distinct() %>%

  # make wider
  #spread(sample_type,sample_value)
  
  pivot_wider(names_from = sample_type,
              values_from = sample_value) %>%
  unnest(cols = c(water_temp_C, air_temp_C, fecal_coliform_cfu_100mL, enterococcus_mpn_100mL, 
    pH, turbidity_NTU, conductivity_mS_cm, wind_velocity_kph, 
    tide_range_m, weather_code_wmo4501, wind_direction_deg)) %>%
  
  distinct() %>%

  # arrange rows chronologically by site
  arrange(site,datetime) 

# print table
datatable(dat,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))
  
```
<br>

Export csv files to local project folder
```{r}

# export csv to local output folder
dir.create("output/virtual_beach_data")

# all overall data
write_csv(dat,"output/vb_data_format.csv",row.names = F)
 
# individual csv for each site
dat %>%
  group_by(site) %>%
  do(write_csv(., paste0("output/virtual_beach_data/",unique(.$site), ".csv")))
```
