---
title: "virtual_beach"
output:
  html_document: 
    fig_width: 10
    fig_height: 6
    fig_caption: yes
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scroll: no
editor_options: 
  chunk_output_type: inline
---

Document last updated `r Sys.time()` by Benjamin Meyer ([ben\@kenaiwatershed.org](mailto:ben@kenaiwatershed.org){.email})

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

# clear environment
rm(list=ls())

# load packages
library(tidyverse)
library(lubridate)
library(readr)
library(readxl)
library(writexl)
library(hms)
library(plotly)
library(DT)
library(xlsx)
library(leaflet)
library(DT)
library(ggpubr)
library(plotrix)
library(zoo)
library(psych)
library(janitor)
library(scales)

select <- dplyr::select

# set plotting themes

## geom_col plots theme
col_theme <- theme(axis.title = element_text(size = 14, face = "bold"),
                   strip.text = element_text(size = 14, face = "bold"),
                   legend.title = element_text(size = 14, face = "bold"),
                   legend.text = element_text(size = 14),
                   axis.text = element_text(size = 14))

## geom_points plots theme
points_theme <- theme(axis.title = element_text(size = 14, face = "bold"),
                   strip.text = element_text(size = 14, face = "bold"),
                   legend.title = element_text(size = 14, face = "bold"),
                   legend.text = element_text(size = 14),
                   axis.text = element_text(size = 14),
                   title = element_text(size = 18))

# function to exclude multiple items per column
'%ni%' <- Negate('%in%')
```

<br>

This draft document prepares and examines data from the EPA BEACON beach database that will be used in implementing the EPA Virtual Beach model for North and South Kenai Beach

<br>

Notes:

-   Data was downloaded as a csv file from EPA BEACON repository using the "reports" function in December 2020 (<https://watersgeo.epa.gov/BEACON2/reports.html>). Contains data 2010 - 2020.

-   Script in development using guidance from virtual beach user manual at <https://www.epa.gov/sites/production/files/2016-03/documents/vb3_manual_3.0.6.pdf>

-   World Meteorological Organization (WMO) 4501 weather codes at <https://www.jodc.go.jp/data_format/weather-code.html>

-   Watch out for mixed use of metric and standard units in original import from EPA database...

-   Need to get associated coordinates for each site

<br>

### Data import

Import and prepare csv file from EPA BEACON database in local project folder

```{r}
# choose variables to retain
vars <- as.character(c("Year",
                       "StationID.Def..",
                       "Station.Name..Def..",
                       "StartDate.Def..",
                       "StartTime.Def..",
                       "ActivityTypeCode.Def..",
                       "CharacteristicName.Def..",
                       "Result.Value.Def..",
                       "Result.MeasureUnit",
                       "Result.Comment.Def..",
                       "ActivityDepthValue.Def.."))

# rename variables
new_vars <- c("year",
              "station_id",
              "site",
              "date",
              "time",
              "sample_activity",
              "sample_type",
              "sample_value",
              "sample_unit",
              "sample_comment",
              "sample_depth_m")


# read in data
dat <- read.csv("data/epa_historical_data.csv") %>%
  select(all_of(vars))

# rename columns
colnames(dat) <- new_vars

# format col type 
dat <- dat %>%
  mutate(datetime = dmy_hms(paste(date,time))) %>%
  transform(date = dmy(date),
            time = hms::as_hms(time),
            sample_value = as.numeric(sample_value),
            sample_type = as.factor(sample_type)) %>%
  # remove missing values
  filter(!is.na(sample_value)) 

# convert all measurement units to METRIC for consistency... sheesh
dat <- dat %>%
  
  # farenheight to celsius
  mutate(sample_value = ifelse(sample_unit == "deg F", (sample_value - 32) / 1.8, sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "deg F", "deg C",sample_unit)) %>%
  
  # miles per hour to kph
  mutate(sample_value = ifelse(sample_unit == "mph",(sample_value*1.609344), sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "mph", "kph",sample_unit)) %>%
  
  # feet to meters
  mutate(sample_value = ifelse(sample_unit == "ft",(sample_value*0.3048), sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "ft", "m",sample_unit)) %>%
  
  # knots to meters per second
  mutate(sample_value = ifelse(sample_unit == "knots",(sample_value*0.51), sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "knots", "mps",sample_unit)) %>%
  
  # inches to meters
  # knots to meters per second
  mutate(sample_value = ifelse(sample_unit == "in",(sample_value*0.0254), sample_value)) %>%
  mutate(sample_unit = ifelse(sample_unit == "in", "m",sample_unit)) 


# modify sample type names
dat <- dat %>%
  mutate(sample_type = str_replace_all(
    sample_type,
    c("Temperature, air" = "air_temp_C",
      "Temperature, water" = "water_temp_C",
      "Fecal Coliform" = "fecal_coliform_cfu_100mL",
      "Enterococcus" = "enterococcus_mpn_100mL",
      "Tide range" = "tide_range_m",
      "Wind velocity" = "wind_velocity_kph",
      "Conductivity" = "conductivity_mS_cm",
      "Turbidity" = "turbidity_NTU",
      "Wave height" = "wave_height_m"
      )))

# some sample type names are stubborn and will not readily modify... may have hidden spaces or other formatting
# use another method
# filter and rename the stubborn sample types
weather <- dat %>%
  filter(sample_type == "Weather condition (WMO code 4501) (choice list)") %>%
  mutate(sample_type = "weather_code_wmo4501")

wind <- dat %>%
  filter(sample_type == "Wind direction (direction from, expressed 0-360 deg)") %>%
  mutate(sample_type = "wind_direction_deg")

# rejoin to overall dataset
dat <- dat %>%
  filter(sample_type %ni% c("Weather condition (WMO code 4501) (choice list)",
                            "Wind direction (direction from, expressed 0-360 deg)")) %>%
  bind_rows(weather,wind)

```

<br>

#### Data summary

```{r}
# summarise
tbl <- dat %>%
  # treat weather codes separately; they are not a continuous variable
  #filter(sample_type != "weather_code_wmo4501") %>%
  group_by(site,sample_type) %>%
  summarise(n = n(),
            mean = round(mean(sample_value),digits = 1),
            min = round(min(sample_value),digits = 1),
            max = round(max(sample_value),digits = 1),
            stderr = round(std.error(sample_value), digits = 1),
            min_year = min(year),
            max_year = max(year))

# print table
datatable(tbl,
          filter = 'top', 
          options = list(pageLength = 8, autoWidth = TRUE))

```

<br>

Notes:

-   There are some observations from 2010 at a site called "Kenai North Beach." This site name nomenclature is different than all the other years. Need to find out which North Kenai Beach site these data are from.
-   WMO Weather codes are not a continuous variable, so mean and std. error do not apply and are spurious values in the above table.

<br>

------------------------------------------------------------------------

#### Replicate samples

Apply criteria for replicate sample values.

-   How many replicate samples do we have?

```{r}
# how many replicate samples do we have?
rep_ct <- dat %>%
  #filter(data_type %in% c("Enterococci","Fecal Coliform","Microbial Source Tracking")) %>%
  group_by(site,date,time,sample_type) %>%
  count(name = "rep_set") %>%
  ungroup() %>%
  filter(rep_set > 1) %>%
  group_by(sample_type,rep_set) %>%
  count(name = "count")

datatable(rep_ct)

  
```

For each sample type, "rep_set" indicates the quantity of simultaneously collected samples in a given instance, and "count" indicates the quantity of instances with simultaneous sample results. For example, as of Jan 15, 2021, we have 81 instances where 2 replicates of "enterococcus_mpn_100mL" exist.

<br>

To employ this data in the Virtual Beach model, we need a single value for each sample type and sampling event.

For environmental variables (temperature, wind, etc), we will create an average value.

For bacteria concentrations however, variation between duplicates can be substantial. ADEC has existing protocol to address this. As per ADEC CALM protocol, "where two replicate values exist, only the higher of the two values are used in further analyses."

In this data set we have additional instances where not just two replicate samples for bacteria concentrations exist, but also instances of 3, 4, 5, and 6 replicates. In these instances also, we will apply the protocol of including only the highest value.

<br>

Apply criteria described above

```{r}
# create list of environmental variables
enviro_vars <- c("wind_velocity_kph",
                 "wave_height_m",
                 "water_temp_C",
                 "tide_range_m",
                 "air_temp_C",
                 "conductivity_mS_cm",
                 "pH",
                 "wind_direction_deg")

# max number of replicate environmental measurements is 2

# create means for environmental variables
enviro_dat <- dat %>%
  filter(sample_type %in% enviro_vars) %>%
  group_by(across(c(-sample_value,-sample_activity,-sample_comment))) %>%
  summarise(mean_sample_value = mean(sample_value)) %>%
  ungroup()

# retain only the highest value in instances where replicate bacteria concentrations values exist
bac_dat <- dat %>%
  filter(sample_type %in% c("fecal_coliform_cfu_100mL","enterococcus_mpn_100mL")) %>%
  group_by(across(c(-sample_value,-sample_activity,-sample_comment))) %>%
  
  # where pairs of two replicate samples exist, retain only the higher of the two values.  
  filter(sample_value == max(sample_value)) %>%
  
  # if two identical replicate values are present, we want to retain only one of the values
  ungroup() %>%
  group_by(across(c(-sample_value,-sample_activity,-sample_comment))) %>%
  summarise(mean_sample_value = mean(sample_value))  

# rejoin bacteria and enviro data
dat <- bind_rows(enviro_dat,bac_dat)

```

<br>

Prep for export

Work in progress here Jan 15, 2021
```{r}
# make environmental variables "wide" format so they each have their own column
dat 

z <- dat %>%

  # remove unneeded columns for clean pivot
  ungroup() %>%
  select(site,date,time,datetime,sample_type,mean_sample_value) %>%
  distinct() %>%
  
  # make wider
  pivot_wider(names_from = sample_type,
              values_from = mean_sample_value) 

# dup values just left for fc 2011 data... why?

z1 <- dat %>%
  filter(sample_)
  
  #%>%
  unnest(cols = c(water_temp_C, air_temp_C, fecal_coliform_cfu_100mL, enterococcus_mpn_100mL, 
    pH, turbidity_NTU, conductivity_mS_cm, wind_velocity_kph, 
    tide_range_m, weather_code_wmo4501, wave_height_m,wind_direction_deg)) %>%
  
  distinct() %>%

  # arrange rows chronologically by site
  arrange(site,datetime) 

# print table
datatable(dat,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))
  
```

<br>

Export csv files to local project folder

```{r}

# export csv to local output folder
dir.create("output/virtual_beach_data")

# all overall data
write_csv(dat,"output/vb_data_format.csv",row.names = F)
 
# individual csv for each site
dat %>%
  group_by(site) %>%
  do(write_csv(., paste0("output/virtual_beach_data/",unique(.$site), ".csv")))
```
